CGAT In The Hat

### 2012-12-07 paper/paper.tex
Made the introduction, motivation, and background awesome. I think. Also added references so that we seem more legit

\section{Introduction}\label{sec:introduction}
Gene annotation is the process of associating metadata about a gene with the
contig, a dna sequence, on which the gene resides.\cite{annotation} This metadata is necessary for conducting
genomic research projects and analysis involving the contig's genomic sequence.
Currently, there is no specialty gene annotation software. Genome browsers
all maintain lots of genomic data and are used when manually annotating genes,
but do not provide a user-friendly interface. While genome browsers
will likely never be replaced (UCSC genome browser, etc. are well
established\cite{ucscbrowser, ncbi}), it would be desirable to have a software system that better
accommodates the visual and informational needs of gene annotation.

\textit{CGAT} is a web-based gene annotation application designed with
usability, simplicity, and efficiency in mind. It is not a feature-rich and data-rich
 genome browser like the \textit{UCSC genome brower}. Nor is \textit{CGAT}
designed to be a replacement for other genome browsers in any aspect other than
gene annotation. Other gene browsers attempt to accommodate many needs of the
biology community and so offer many features, and a cluttered, chaotic
interface. Gene annotation, while not a trivial task, can be well-serviced by a
simple data model. Ideally, by focusing on a simple data model, \textit{CGAT}
is able to provide a clean, streamlined experience. \textit{CGAT} aims to
provide users with a way to view gene annotations without extra, unnecessary
information obscuring their view while also embodying a Wikipedia-like emphasis
on collaboration and openness.

This paper discusses the progress that has been made this quarter and the future work that 
still needs to be done. The rest of the paper is as follows. Section \ref{sec:background} discusses
the common terms and biological meanings used in the project. Section \ref{sec:motive} discusses the
motivation for creating CGAT. Section \ref{sec:choice} explains why we chose MongoDB
as our backend. Section \ref{sec:implementation} goes into the details of how
our backend and frontend are implemented currently. Section \ref{sec:results} discusses
the results of our demo to our customer. Lastly, section \ref{sec:future} 
describes the future work still left to do.

\section{Background}\label{sec:background}

\section{Motivation}\label{sec:motive}
Dr. Anya Goodman, a professor in the Chemistry department at Cal Poly, San Luis
Obispo, offers a bioinformatics course that covers several aspects of gene
annotation. Her work is a part of the Genomics Education Partnership (GEP) at 
Washington University in St. Louis.\cite{gep} The goal of GEP is to teach undergrad students
in Biology how to annotate genes and other features in raw sequence data.
The CGAT project is spearheaded by Ryan Verdon to accommodate Dr.
Goodman's needs and ideas for ideal gene annotation software.

>>>>

% TODO include more background:
% http://www.weizmann.ac.il/complex/tlusty/courses/landmark/JacobMonod1960.pdf
% http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2443188/
% http://www.sciencemag.org/content/suppl/2012/04/11/336.6078.179.DC1/Zerbino.SM.pdf
% http://gep.wustl.edu/community/publications
% http://gep.wustl.edu/projects/dot_chromosome
\section{Introduction}\label{sec:introduction}
Dr. Anya Goodman, a professor in the department of chemistry and biochemistry
at Cal Poly, San Luis Obispo, offers a bioinformatics course that covers
several aspects of gene annotation and genomic research. Her work is a part of
the \textit{Genomics Education Partnership} (GEP) which is based at Washington
University in St. Louis~\cite{gep}. GEP is an organization that aims to bring
genomics research to academic institutions, especially undergraduate classes.
This allows undergraduate students to get the opportunity to learn about, and
practice, basic genomic research and, in turn, contribute meaningful work to
the biological research community. However, GEP does not have an existing
system in which users can input and save gene annotations and get immediate
feedback regarding their performance. Similarly, available gene annotation
software--RepeatMasker, Artemis, and others~\cite{repeatmasker, artemis}--is
not designed to provide immediate feedback for users regarding their
performance. In this paper, we present Community Genome Annotation Trainer,
\textit{CGAT}, a web-based gene annotation application designed to provide
feedback of gene annotation quality to students faster.

\textit{CGAT} is spearheaded by Ryan Verdon and has been built with usability,
simplicity, and efficiency in mind. It is not a feature-rich and data-rich
genome browser like \textit{UCSC genome brower}~\cite{ucscbrowser},
\textit{Ensembl}~\cite{ensembl}, or \textit{BioViz}~\cite{bioviz}, nor is it
designed to be a replacement for gene annotatioin software such as
RepeatMasker, Artemis, or any others, in any aspect other than for teaching, or
practicing, gene annotation. Other gene browsers or annotation software offer
many features to accommodate the needs of the biology community, and so tend to
be complex. Gene annotation, while not a trivial task, can be well-serviced by
a simple data model. Ideally, by focusing on a simple data model, \textit{CGAT}
is able to provide a clean, streamlined experience capable of providing users
with a way to view gene annotations made by others or have their gene
annotations reviewed by experts. In particular, \textit{CGAT} strives towards a
Wikipedia-like emphasis on collaboration and openness.

This paper discusses the progress that has been made during a graduate database
class at Cal Poly, San Luis Obispo on \textit{CGAT} as well as relevant future
work. The rest of the paper is as follows. Section \ref{sec:background}
discusses the common terms and biological meanings used in the project. Section
\ref{sec:motive} discusses the motivation for creating CGAT. Section
\ref{sec:choice} explains why we chose MongoDB as our backend. Section
\ref{sec:implementation} goes into the details of how our backend and frontend
are implemented currently. Section \ref{sec:results} discusses the results of
our demo to our customer. Lastly, section \ref{sec:future} describes the future
work still left to do.

\section{Background}\label{sec:background}
Genomics research attempts to understand information contained in the
\textit{genomes} of various organisms. The genome is the entire DNA sequence
that encodes information and drives all cellular processes in an organism.
However, due to the size of genomes, biologists are only able to sequence
sections, or fragments, of a genome at a time. One or more contiguous fragments
of the genome together form a \textit{contig}, or simply a contiguous sequence
of DNA nucleotides. The genome contains several components, or sub-regions,
that have different functions or roles. These components describe in detail how
a genome is decoded to drive cellular processes. \textit{Annotations}--the
process of linking biological information of the organism to the genome--are an
important step to being able to relate information learned from the sequence back
to broader understandings of the associated organism and, ideally, related
organisms~\cite{annotation}. Again, since the genome is so large, biologists
typically annotate sections of the genome, specifically \textit{genes}. A gene
is a section of the genome that codes for a functional \textit{protein} used
for biological processes inside and outside of the cell. A gene contains
several components, or sub-regions, that have different functions or
roles--\textit{exons}, \textit{introns}, \textit{CDS}s, \textit{UTR}s, and
others. These components that are located in a gene some of the most basic but
most important pieces of information that must be annotated in order to be able
to make meaningful biological interpretations from information learned from the
genomic sequence, and where our project, \textit{CGAT}, is useful. Considering
the scope of an undergraduate genomics research class, it is preferable to have
a system that emphasizes simplicity and usability as well as houses information
in a collaborative and educational environment.
